---
title: "Influenza Antibody Analysis Report"
author: "Lukas Baumann, 14.04.2025"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    warning: false
    echo: false
execute:
  cache: false   
editor: visual
---

# HAI Analysis

```{r}
#| include: false
source(here::here("scripts", "02_hai_microneut_basic_analysis.R"))
```

**Display pre-post vaccine HAI-titers by group and strain**

```{r}
plot_hai_comb_group_wide
```

# Microneutralisation Titers

**Display pre-post vaccine microneutralisation titers by group and strain**

```{r}
plot_ic50_comb_group_wide
```

# Correlation of Baseline Titers with Fold-Changes

```{r}
#| include: false
source(here::here("scripts", "03_baseline_titers_assessment.R"))
```

## Different linear regression models to assess optimal way to adjust for basline titers

1)  log2 Fold-Changes (y-axis) and untransformed vs. log2-transformed baseline titers

```{r}
linreg_plot_comb
```

2)  log2 Fold-Changes vs. log2 first Fold-Changes (y-axis) and log2-transformed baseline titers

```{r}
linreg_plot_comb2
```

## Comparing correlation of Microneutralisation Fold Changes to baseline titers with different models

**H1N1 Strain**

```{r}
cor_comb1
```

**H3N2 Strain**

```{r}
corh3_comb
```

**B Victoria Strain**

```{r}
corb_comb
```

--\> base_log2 vs. log2 first has the highest correlation coefficient for all 3 strains. log2 or log10 gives same results.

# Linear Regression Models & Residual Diagnostics

```{r}
linreg_plot_test_combined
```

# Correlation of HAI and Microneutralisation Assays

```{r}
#| include: false
source(here::here("scripts", "04_hai_vs_microneut.R"))
```

```{r}
ggplot(data=df3, aes(x=hai_rank, y=ic50_rank)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  annotate("text",
           x = Inf, y = -Inf,
           hjust = 1.1, vjust = -0.7,
           label = paste0("Spearman ρ = ", rho, "\nP = ", pval),
           size = 4) +
  labs(title = "Rank Correlation: HAI vs IC50",
       x = "HAI Rank",
       y = "IC50 Rank") +
  theme_minimal()
```
