---
title: "FluVacc_HAI_vs_Microneut"
author: "Lukas Baumann" 
date: 2025-06-03
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo 
    self-contained: true
    fontsize: 1.1em
    linestretch: 1.7
    number-sections: false
    code-fold: true
    code-summary: "Show Code"
    warning: false
    echo: false
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    echo: false
    warning: false
    keep-tex: true
    documentclass: article
    geometry: margin=1in
execute:
  cache: false   
editor: visual
---

# HAI Titers Overview

```{r}
#| include: false
source(here::here("scripts", "02_hai_microneut_basic_analysis.R"))
```

**Display pre-post vaccine HAI-titers by group and strain**

```{r}
plot_hai_comb_group_wide
```

# Microneutralisation Titers Overview

**Display pre-post vaccine microneutralisation titers by group and strain**

```{r}
plot_ic50_comb_group_wide
```

```{r}
#| include: false
source(here::here("scripts", "04_hai_vs_microneut.R"))
```

# Box-plots of uncorrected fold change values

-   y axis is clipped at 50 (outliers above are not shown)
-   horizontal lines at Fold-changes of 1(red), 2(orange) and 4(green)

```{r}
influenza_folds_decline %>% 
  select(strain, assay, fold_change) %>%  # use `value` from pivoted data, not `fold_change`
  #filter(fold_change < 1) %>%   # <- Filter values below 1
  #filter(strain != "B/Yam") %>% 
  mutate(strain = as.factor(strain)) %>% 
  ggplot(aes(x = strain, y = fold_change, color = strain)) +
  geom_boxplot() +
  #geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed") +  # <- red line at y = 1
  geom_hline(yintercept = 2, color = "orange", linetype = "dashed") + 
  geom_hline(yintercept = 4, color = "green", linetype = "dashed") +  
  facet_wrap(~ assay) +
  coord_cartesian(ylim = c(0, 50)) +  # limit visible y-axis range
  theme_minimal()
```

# Correlation of HAI and Microneutralisation Assays

```{r}
#| warning: false
hai_vs_ic50_ranksumplot
```

```{r}
strain_corrs_tidy
```

**Interpretation**: Moderate - strong association between HAI and Microneutralisation assays. Lower for H3N2.

# Decline in titers (FoldChange \< 1)

```{r}
influenza_folds_decline %>% 
  count(assay, decline) %>% 
  group_by(assay) %>% 
  mutate(percentage = n / sum(n) * 100) %>%  
  print()
```

--\> 26.9% (Microneutralisation) vs 4.5% (HAI) with declining titers - Unclear

```{r}
influenza_folds_decline %>% 
  select(strain, assay, fold_change) %>%  # use `value` from pivoted data, not `fold_change`
  filter(fold_change < 1) %>%   # <- Filter values below 1
  filter(strain != "B/Yam") %>% 
  mutate(strain = as.factor(strain)) %>% 
  ggplot(aes(x = strain, y = fold_change, color = strain)) +
  geom_boxplot() +
  geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5) +
  facet_wrap(~ assay) +
  theme_minimal()
```

Baseline Titer in Patients with declining Fold Changes

```{r}
influenza_records_declining_titers %>% 
  filter(assay == "mic") %>% 
  ggplot(aes(x = decline, y = value, color = decline)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    x = "Fold-Change Decline",
    y = "Microneutralisation Baseline Titer"
  )
```

# Response per strain

## Response overview

```{r}
count_table_response2_percent %>% 
  filter(strain != "B/Yamagata") %>% 
  ggplot() +
  geom_col(aes(x = strain, y = percentage, fill = assay),
           position = position_dodge()) +
  geom_text(aes(x = strain, y = percentage, label = round(percentage, 1), group = assay),
            position = position_dodge(width = 0.9),
            vjust = -0.3, size = 3) +
  labs(x = "Strain", 
       y = "Percentage", 
       fill = "Assay") +
  facet_wrap(~ fold_category) +
  theme_minimal()
```

##HAI

```{r}
ggplot(data = count_table_response2 %>% filter(assay == "hai")) + 
  geom_col(aes(y = count, x = strain, fill = fold_category), position = "dodge") +
  scale_fill_manual(
    values = c(
      "low response" = "red",
      "moderate response" = "orange",
      "high response" = "green"
    )
  ) +
  labs(title = "HAI Response by Strain",
       x = "Strain",
       y = "Count",
       fill = "Response") +
  theme_minimal()
```

in percent:

```{r}
# Plot percentages (filtered to HAI, optional)
ggplot(data = count_table_response2_percent %>% filter(assay == "hai")) +
  geom_col(aes(x = strain, y = percentage, fill = fold_category), position = "stack") +
  scale_fill_manual(
    values = c(
      "low response" = "red",
      "moderate response" = "orange",
      "high response" = "green"
    )
  ) +
  labs(title = "HAI Response by Strain (as %)",
       x = "Strain",
       y = "Percentage",
       fill = "Response Category") +
  theme_minimal()
```

## Microneutralisation

```{r}
ggplot(data = count_table_response2 %>% filter(assay == "mic")) + 
  geom_col(aes(y = count, x = strain, fill = fold_category), position = "dodge") +
  scale_fill_manual(
    values = c(
      "low response" = "red",
      "moderate response" = "orange",
      "high response" = "green"
    )
  ) +
  labs(title = "Microneutralisation Response by Strain",
       x = "Strain",
       y = "Count",
       fill = "Response") +
  theme_minimal()
```

in percent

```{r}
# Plot percentages (filtered to mic, optional)
ggplot(data = count_table_response2_percent %>% filter(assay == "mic")) +
  geom_col(aes(x = strain, y = percentage, fill = fold_category), position = "stack") +
  scale_fill_manual(
    values = c(
      "low response" = "red",
      "moderate response" = "orange",
      "high response" = "green"
    )
  ) +
  labs(title = "Microneutralisation Response by Strain (as %)",
       x = "Strain",
       y = "Percentage",
       fill = "Response Category") +
  theme_minimal()
```

-\> BVictoria and H1N1 with slightly lower response rates

# Heatmaps HAI vs. Microneutralisation

-   Categorised by unadjusted titer fold changes and sorted by approximate overall fold change across strains:

-   Categories

    -   Red = low response = "\<2-fold"

    -   Orange = moderate response = "2–4-fold"

    -   Green = high response = "\> 4-fold"

## Overall Heatmap

```{r, fig.width=12, fig.height=25}
ggplot(influenza_folds_custom, aes(x = x, y = Pat_ID, fill = Fold_Category)) +
  geom_tile(color = "white", width = 0.6) +
  scale_fill_manual(
    values = c("1" = "red", "3" = "orange", "4" = "green", "99" = "grey"),
    name = "Fold Category",
    labels = c("1" = "<2-fold", "3" = "2–4-fold", "4" = ">4-fold", "99" = "Other")
  ) +
  scale_x_continuous(
    breaks = as.numeric(names(x_label_map)),
    labels = x_label_map
  ) +
  labs(x = "Assay Type", y = "Patient ID") +
  coord_fixed(ratio = 0.3) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

```

-\> Visually acceptable overlapp in response across strains and assay type with some exceptions.

## Faceted by patient group

```{r, fig.width=12, fig.height=20}
ggplot(influenza_folds_custom, aes(x = x, y = Pat_ID, fill = Fold_Category)) +
  geom_tile(color = "white", width = 0.6) +
  scale_fill_manual(
    values = c("1" = "red", "3" = "orange", "4" = "green", "99" = "grey"),
    name = "Fold Category",
    labels = c("1" = "<2-fold", "3" = "2–4-fold", "4" = ">4-fold", "99" = "Other")
  ) +
  scale_x_continuous(
    breaks = as.numeric(names(x_label_map)),
    labels = x_label_map
  ) +
  labs(
    x = "Assay Type", y = "Patient ID"
  ) +
  facet_wrap(~ pat_group, scales = "free_y", ncol = 2) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold")
  )
```

-\> more low-responders in Onco-group, otherwise visually not clearly different

# Overal response across different strains

## Microneutralisation

```{r}
influenza_overall %>%  count(overall_response_micr) %>% 
  print()
```

## HAI

```{r}
influenza_overall %>%  count(overall_response_hai) %>% 
  print()
```

```{r, fig.width=12, fig.height=20}
# Reshape to long format for heatmap
df_long <- influenza_overall %>%
  pivot_longer(cols = starts_with("overall_response"), 
               names_to = "assay", 
               names_prefix = "overall_response_",
               values_to = "response")

# Define color category and Add descriptive label mapping
df_long <- df_long %>%
  mutate(response_label = case_when(
    response == "all high" ~ "All strains high response",
    response == "all low" ~ "All strains low response",
    TRUE ~ "Mixed response"
  ))

# Create the faceted heatmap
ggplot(df_long, aes(x = assay, y = Pat_ID, fill = response_label)) +
  geom_tile(color = "white") +
  scale_fill_manual(
    values = c(
      "All strains high response" = "green",
      "Mixed response" = "orange",
      "All strains low response" = "red"
    )
  ) +
  facet_wrap(~ pat_group, scales = "free_y") +  # Facet by patient group
  theme_minimal() +
  labs(title = "Overall response across strains - Heatmap by Group",
       x = "Assay Type",
       y = "Patient ID",
       fill = "Response Category") + 
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    strip.text = element_text(face = "bold")
  )
```

-\> only few patients with only high- or low response across strains. BUT: criterion applied is very strict, as truly all strain-responses need to be \>4 or \<2 to count as "all strains".
