---
title: "Influenza Antibody Analysis Report"
author: "Lukas Baumann, 14.04.2025"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    warning: false
    echo: false
execute:
  cache: false   
editor: visual
---

# HAI Analysis

```{r}
#| include: false
source(here::here("scripts", "02_hai_microneut_basic_analysis.R"))
```

**Display pre-post vaccine HAI-titers by group and strain**

```{r}
plot_hai_comb_group_wide
```

# Microneutralisation Titers

**Display pre-post vaccine microneutralisation titers by group and strain**

```{r}
plot_ic50_comb_group_wide
```

# Correlation of Microneutralisation Baseline Titers with Fold-Changes

```{r}
#| include: false
source(here::here("scripts", "03_baseline_titers_assessment.R"))
```

## Different linear regression models to assess optimal way to adjust for basline titers

### log2 Fold-Changes (y-axis) and untransformed vs. log2-transformed baseline titers

```{r}
linreg_plot_comb
```

**Interpretation**: log2 transformed baseline titers leading to better fitting linear models

### log2 Fold-Changes vs. log2 first Fold-Changes (y-axis) and log2-transformed baseline titers

```{r}
linreg_plot_comb2
```

**Interpretation**: "log2 first transformed" Fold-changes leading to better fitting linear models

## Comparing correlation of Microneutralisation Fold Changes to baseline titers with used different models

**H1N1 Strain**

```{r}
cor_comb1
```

**H3N2 Strain**

```{r}
corh3_comb
```

**B Victoria Strain**

```{r}
corb_comb
```

**Interpretation:** model with base_log2 vs. log2first fold changes has the highest correlation coefficient for all 3 strains. log2 or log10 gives same results. Seems most appropriate to use log2 first fold-change models for linear regressions.

# Assessing Linearity Asumption for Regression Models (FoldChange - baselineTiters)

## QQ Plots plots from univariant regression models for all strains - comparing fold-changes log2 and log2 first FC

```{r}
sumplot_qq_uni
```

**Interpretation**: Visually ok-ish looking qq-plots. Some divergence from assumed normal-distribution for lowest- and highest values. No relevant difference between log2 and log2first models.

## Check for homoscedasticity using residuals vs fitted values plot

```{r}
residplot_combined
```

**Interpretation**: visually no relevant concearns about heteroscedasticity in the regression models. Log2 vs. Log2 first noch visually different.

##Look at autocorrelation (Durbin Whatson Test) and heteroscedasticity (Beusch-pagan test)

```{r}
bp_dw_results
```

**Interpretation**:

-   Relevant (high-moderate) Heteroscedasticity, with relevant impact of log2 first approach in the case of H1N1 (worse)

-   some Autocorrelation, but not hugely bothering. No relevant impact of log2 vs. log2first

## Compare Adjusted R2, AIC and Estimates of models (log2 vs. log2 first Fold Change)

```{r}
model_results_nice
```

**Interpretation**:

-   Log2 FC models show larger (absolute) effect sizes, but poorer model fit and much worse AIC. But log2first seem not sensitivie to do as it is unclear how these values can be backtransformed(ToClarify).

-   Log2-first models reduce the effect size (see above), but have improved model fit (R²) and model quality (AIC).

# Multivariant models for microneutralisation fold-changes
foldchange vs. log2 baseline model
```{r}
sum_b_model2
```
```{r}
#| include: false
source(here::here("scripts", "05_micneut_basecorrect_beyer04.R"))
```
log2 corrected_outcome titers vs. log2 baseline model
```{r}
sum_b_model4
```

# Correction of outcome titers by linear regression according to beyer et.al. 2004

-   Formula:\
    outcome_titers_corrected = log2(outcome_titers) - slope \* (log2(baseline_titers) - baseline_constant)

-   Where:

    -   baseline_constant = baseline_value to which outcomes are corrected (log2(39) for seronegative)

    -   slope: taken from linear regression model

        -   lm(log2(outcome_titers) \~ log2(baseline_titers))
        
```{r}
#| include: false
source(here::here("scripts", "05_micneut_basecorrect_beyer04.R"))
```

```{r}
b_titers_uncorrected
```
```{r}
b_titers_corrected
```

# Correlation of HAI and Microneutralisation Assays

```{r}
#| include: false
source(here::here("scripts", "04_hai_vs_microneut.R"))
```

```{r}
hai_vs_ic50_ranksumplot
```

**Interpretation**: Moderate - strong association between HAI and Microneutralisation assays with an spearman rho of 0.69. (CAVE: ranking only within the same strains, but result not much different if done otherwise - rho 0.66)
