---
title: "Influenza Antibody Analysis Report"
author: "Lukas Baumann, 14.04.2025"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    warning: false
    echo: false
execute:
  cache: false   
editor: visual
---

# HAI Analysis

```{r}
#| include: false
source(here::here("scripts", "02_hai_microneut_basic_analysis.R"))
```

**Display pre-post vaccine HAI-titers by group and strain**

```{r}
plot_hai_comb_group_wide
```

# Microneutralisation Titers

**Display pre-post vaccine microneutralisation titers by group and strain**

```{r}
plot_ic50_comb_group_wide
```

# Correlation of Microneutralisation Baseline Titers with Fold-Changes

```{r}
#| include: false
source(here::here("scripts", "03_baseline_titers_assessment.R"))
```

## Different linear regression models to assess optimal way to adjust for basline titers

### log2 Fold-Changes (y-axis) and untransformed vs. log2-transformed baseline titers

```{r}
linreg_plot_comb
```

**Interpretation**: log2 transformed baseline titers leading to better fitting linear models

### log2 Fold-Changes vs. log2 first Fold-Changes (y-axis) and log2-transformed baseline titers

```{r}
linreg_plot_comb2
```

**Interpretation**: "log2 first transformed" Fold-changes leading to better fitting linear models

## Comparing correlation of Microneutralisation Fold Changes to baseline titers with used different models

**H1N1 Strain**

```{r}
cor_comb1
```

**H3N2 Strain**

```{r}
corh3_comb
```

**B Victoria Strain**

```{r}
corb_comb
```

**Interpretation:** model with base_log2 vs. log2first fold changes has the highest correlation coefficient for all 3 strains. log2 or log10 gives same results. Seems most appropriate to use log2 first fold-change models for linear regressions.

# Assessing Linearity Asumption for Regression Models

## QQ Plots plots from univariant regression models for all strains - comparing fold-changes log2 and log2 first

```{r}
sumplot_qq_uni
```

**Interpretation**: Visually ok-ish looking qq-plots. Some divergence from assumed normal-distribution for lowest- and highest values. No relevant difference between log2 and log2first models.

## Check for homoscedasticity using residuals vs fitted values plot

```{r}
residplot_combined
```

**Interpretation**: visually no relevant concearns about heteroscedasticity in the regression models. Log2 vs. Log2 first noch visually different.

##Look at autocorrelation (Durbin Whatson Test) and heteroscedasticity (Beusch-pagan test)

```{r}
bp_dw_results
```

**Interpretation**:

-   Relevant (high-moderate) Heteroscedasticity, with relevant impact of log2 first approach in the case of H1N1 (worse)

-   some Autocorrelation, but not hugely bothering. No relevant impact of log2 vs. log2first

## Compare Adjusted R2, AIC and Estimates of models (log2 vs. log2 first Fold Change)

```{r}
model_results_nice
```

**Interpretation**:

-   Raw models show larger (absolute) effect sizes, but poorer model fit and much worse AIC. To clarify how log2 transformed Fold changes should be backtransformed (log2 vs. log2 first) - likely most of the difference in effect sizes are due to the way fold-changes are log-transformed.

-   Log2-first models reduce the effect size (see above), but greatly improve model fit (R²) and model quality (AIC).

# Correlation of HAI and Microneutralisation Assays
```{r}
#| include: false
source(here::here("scripts", "04_hai_vs_microneut.R"))
```

```{r}
hai_vs_ic50_ranksumplot
```
**Interpretation**: Moderate - strong association between HAI and Microneutralisation assays with an spearman rho of 0.69. (CAVE: ranking only within the same strains, but result not much different if done otherwise - rho 0.66)