---
title: "Influenza Antibody Analysis Report"
author: "Lukas Baumann, 14.04.2025"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    warning: false
    echo: false
execute:
  cache: false   
editor: visual
---

# HAI Analysis

```{r}
#| include: false
source(here::here("scripts","02_hai_microneut_basic_analysis.R"))
```

**Display pre-post vaccine HAI-titers by group and strain**

```{r}
plot_hai_comb_group_wide
```

# Microneutralisation Titers

**Display pre-post vaccine microneutralisation titers by group and strain**

```{r}
plot_ic50_comb_group_wide
```

# Correlation of Microneutralisation Baseline Titers with Fold-Changes

```{r}
#| include: false
source(here::here("scripts", "03_baseline_titers_assessment.R"))
```

## Different linear regression models to assess optimal way to adjust for basline titers

### log2 Fold-Changes (y-axis) and untransformed vs. log2-transformed baseline titers

```{r}
linreg_plot_comb
```

**Interpretation**: log2 transformed baseline titers leading to better fitting linear models

### log2 Fold-Changes vs. log2 first Fold-Changes (y-axis) and log2-transformed baseline titers

```{r}
linreg_plot_comb2
```

**Interpretation**: "log2 first transformed" Fold-changes leading to better fitting linear models

## Comparing correlation of Microneutralisation Fold Changes to baseline titers with used different models

**H1N1 Strain**

```{r}
cor_comb1
```

**H3N2 Strain**

```{r}
corh3_comb
```

**B Victoria Strain**

```{r}
corb_comb
```

**Interpretation:** model with base_log2 vs. log2first fold changes has the highest correlation coefficient for all 3 strains. log2 or log10 gives same results. Seems most appropriate to use log2 first fold-change models for linear regressions.

# Assessing Linearity Asumption for Regression Models (FoldChange - baselineTiters)

## QQ Plots plots from univariant regression models for all strains - comparing fold-changes log2 and log2 first FC

```{r}
sumplot_qq_uni
```

**Interpretation**: Visually ok-ish looking qq-plots. Some divergence from assumed normal-distribution for lowest- and highest values. No relevant difference between log2 and log2first models.

## Check for homoscedasticity using residuals vs fitted values plot

```{r}
residplot_combined
```

**Interpretation**: visually no relevant concearns about heteroscedasticity in the regression models. Log2 vs. Log2 first noch visually different.

##Look at autocorrelation (Durbin Whatson Test) and heteroscedasticity (Beusch-pagan test)

```{r}
bp_dw_results
```

**Interpretation**:

-   Relevant (high-moderate) Heteroscedasticity, with relevant impact of log2 first approach in the case of H1N1 (worse)

-   some Autocorrelation, but not hugely bothering. No relevant impact of log2 vs. log2first

## Compare Adjusted R2, AIC and Estimates of models (log2 vs. log2 first Fold Change)

```{r}
model_results_nice
```

**Interpretation**:

-   Log2 FC models show larger (absolute) effect sizes, but poorer model fit and much worse AIC. But log2first seem not sensitivie to do as it is unclear how these values can be backtransformed(ToClarify).

-   Log2-first models reduce the effect size (see above), but have improved model fit (R²) and model quality (AIC).

# Correction of outcome titers by linear regression according to beyer et.al. 2004

-   Formula:\
    outcome_titers_corrected = log2(outcome_titers) - slope \* (log2(baseline_titers) - baseline_constant)

-   Where:

    -   baseline_constant = baseline_value to which outcomes are corrected (log2(39) for seronegative)

    -   slope: taken from linear regression model

        -   lm(log2(outcome_titers) \~ log2(baseline_titers))

```{r}
#| include: false
source(here::here("scripts", "05_micneut_basecorrect_beyer04.R"))
```

```{r}
b_titers_uncorrected
```

```{r}
b_titers_corrected
```

# Multivariant models for microneutralisation fold-changes

```{r}
#| include: false
source(here::here("scripts", "03_baseline_titers_assessment.R"))
```

foldchange vs. log2 baseline model

```{r}
sum_b_model2
```

```{r}
#| include: false
source(here::here("scripts", "05_micneut_basecorrect_beyer04.R"))
```

log2 corrected_outcome titers vs. log2 baseline model

```{r}
sum_b_model4
```

# Fold Changes: Different Calculations

-   Hot to adjust for baseline differences:

    -   Stratification by similar baseline titers

    -   Regression / Analysis of Covariance

-   Different Outcome measures

    -   Max Titer vs Sum Titer response

    -   Category of Responses (30/70 or 20/80% cut-off for low-/medium-/high responders)

    -   Absolute response categories:

        -   seroprotection rate (for HAI \>1:40)

        -   unadjusted response rate (Fold increase ≥4x)

1\) Global Analyses of Human Immune Variation Reveal Baseline Predictors of Postvaccination Responses, Cell 2014

<http://dx.doi.org/10.1016/j.cell.2014.03.031>

-   maximum FC across viral strains: To take the maximum of individual viruses, we performed standardization by subtracting the median followed by dividing the maximum absolute deviation (MAD) within each virus.

-   adjusted FC: binned subjects based on their maximum baseline titers (across viruses) and then

    -   subtracted the median and

    -   divided by the maximum absolute deviation (MAD) within each bin.

-   high- and low responders (top and bottom 20%)

-   Ref:

    -   Nauta, J. (2011). Statistics in clinical vaccine trials (Heidelberg, Germany: Springer).

    -   Bucasas, K.L., Franco, L.M., Shaw, C.A., Bray, M.S., Wells, J.M., Nin˜o, D., Arden, N., Quarles, J.M., Couch, R.B., and Belmont, J.W. (2011). Early patterns of gene expression correlate with the humoral immune response to inﬂuenza vaccination in humans. J. Infect. Dis. 203, 921–929.

2\) Multicohort analysis reveals baseline transcriptional predictors of influenza vaccination responses, Science Immunology 2017

doi:10.1126/sciimmunol.aal4656.

-   adjusted maximum fold change” (adjMFC):

    -   To help ensure that baseline titers and response fold changes of individual strains were comparable, so that the maximum is meaningful, we standardized titers for each strain at baseline (day 0) by subtracting the median and then scaling by the SD.

    -   We then defined the baseline (relative) titer for each individual as the maximum across standardized baseline values for all strains.

    -   Similarly, we used the same approach to standardize the titer response fold change (usually day 28 over day 0) and called these the MFC. In addition, we applied the inverse normal transformation to the MFC values to avoid extreme skews in the distribution. We next plotted baseline and MFC, and as previously observed, we saw a strong nonlinear inverse correlation between them in all of our data sets. To compute adjMFC (i.e., removing the nonlinear correlations), we binned the individuals based on their baseline titer so that the correlation between baseline and response titers within each bin was not statistically significant, and at the same time, we tried to keep the number of bins as low as possible so that each bin would have a reasonable number of individuals for decorrelation. Typically, two to three bins were used. In some of our cohorts, there were a few outlier individuals with extremely high baseline titers and correspondingly small response fold changes; they were removed from further analysis. For each bin, we computed the decor-related responses (adjMFC) by standardizing the MFC values within each bin, that is, subtracting the median and scaling by the SD. Last, individuals within each cohort were defined as low, moderate, and high responders using percentile cutoffs (7). To accommodate the relatively small number of individuals in some of the cohorts, we defined low, moderate, and high responders as those whose adjMFC was lower than the 30th percentile value, between the 30th and the 70th percentile values, and above the 70th percentile value, respectively. After applying the decorrelation procedure, we also performed a number of diagnostic checks to ensure that our approach had achieved the desired effects. Specifically, we confirmed the following: (i) The correlation between baseline titer and the adjMFC was removed on the basis of the Spearman rank correlation; (ii) at the individual strain level, the correlation between the baseline titer and the response fold change was statistically insignificant (i.e., before aggregation of the individual titer per strain via computing the maximum); (iii) the identity of the strain that contributed to the baseline (or response fold change) maximum did not correlate with adjMFC. In all of the above checks, two variables were considered not correlated if the Spearman correlation P value was greater than 0.1.

# Open Questions

1)  Is there a way to backtransform log2first FC?
